\begin{abstractCN}
\setlength{\baselineskip}{23pt} %设置行距23磅

随着物联网的发展，网络中的状态更新业务不断涌现，并广泛应用于工业物联网、车路协同等典型场景中。对这类业务而言，接收端更关注状态信息的新鲜度，以期为后续的系统监测、状态估计与控制决策提供可靠支撑。传统的吞吐量和时延指标难以对信息的新鲜度进行衡量，为解决这个问题，信息年龄（Age of Information，AoI）这一新兴指标被提出并受到广泛关注。

随着物联网部署规模的持续扩张，状态更新业务的设备数量呈现出海量化的演进趋势。在这种大规模接入的场景下，若持续依赖集中式调度，系统需要维护大量节点的状态，这将承担极高的控制和反馈开销。相比之下，随机接入则避免了对全局状态信息的掌控，能够在低复杂度的情况下支持大规模设备的动态接入，在拥有海量设备的通信场景中更加适用。正因如此，如何在大规模接入的条件下兼顾接入效率与信息新鲜度，是一个值得重点研究的 MAC 层问题。许多学者围绕这一问题开展研究，使得这一领域涌现诸多成果。

然而，现有研究没有充分考虑大规模接入场景下持续碰撞所带来的影响，而且大部分研究的讨论重点都放在平均 AoI 上，对 AoI 违规概率这类尾部指标的关注较少。对于工业控制等时间敏感型业务，仅考察平均 AoI 不足以判断系统的可靠性。针对上述问题，本文充分考虑了大规模接入场景下持续碰撞所带来的负面影响，先是设计了一种基于等待的随机接入退避机制，并在此基础上，围绕时隙 ALOHA 和标准 CSMA 两类经典的随机接入框架来构建相应的理论分析模型，推导出了平均 AoI 与 AoI 违规概率的闭式表达式。本文的主要工作如下：

\begin{enumerate} [wide]
    \item \textbf{设计面向大规模接入场景的等待机制。}针对现有研究未充分考虑大规模接入场景下持续碰撞所带来的影响这一问题，本文设计了一种基于等待的随机接入退避机制，该机制将碰撞行为纳入考虑范围，通过规划节点在成功更新以及遭遇碰撞后的不同退避行为，有效缓解了信道的过度竞争，从而弥补了现有机制在拥塞处理方面的空缺。
    \item \textbf{将所提机制引入时隙 ALOHA 框架中并进行性能分析。} 针对大规模网络中节点状态耦合、系统难以解析的难题，本文通过稳态解耦的近似方法，将高维复杂的全局状态降维成单节点的马尔可夫链模型，在此基础上建立了系统的稳态方程并对其求解。随后，基于求得的稳态特征，推导出平均 AoI 及 AoI 违规概率的闭式表达式，并深入分析了关键的系统参数对信息时效性的影响。
    \item \textbf{将所提机制引入标准 CSMA 框架中并进行性能分析。} 针对大规模网络中节点状态耦合与 CSMA 框架中载波侦听、退避冻结等机制导致的系统时隙变长两个难题，本文通过稳态解耦的近似方法，将高维复杂的全局状态降维成单节点的马尔可夫链模型。在考虑变长时隙的前提下，基于马尔可夫链模型建立了系统的稳态方程并对其求解。随后，将求得的稳态特征与变长时隙的统计特性相结合，推导出平均 AoI 及 AoI 违规概率的闭式表达式，并深入分析了关键的系统参数对信息时效性的影响。
    \item \textbf{验证理论模型并揭示等待机制的性能演化规律。}通过仿真实验，充分验证了本文所推导的闭式表达式的准确性。进一步的对比分析表明，在中高负载尤其是海量设备接入的场景下，适度等待并不会损害信息新鲜度。相反，它能够减少信道中的竞争，降低碰撞概率，从而同时改善平均 AoI 与 AoI 违规概率。
\end{enumerate}

综上所述，本文的研究不仅丰富了面向信息年龄的随机接入理论分析体系，也为未来大规模物联网状态更新系统的协议设计与优化提供了有价值的理论参考与工程指导。

\end{abstractCN}


\begin{keywordCN}
随机接入，信息年龄，性能分析
\end{keywordCN}

\begin{abstractEN}
\setlength{\baselineskip}{23pt} %设置行距23磅

With the development of the Internet of Things (IoT), status update services continue to emerge in networks. They are widely used in typical scenarios such as the Industrial IoT and vehicle-to-infrastructure cooperation. For these services, the receiver cares more about the freshness of status information. This provides reliable support for subsequent system monitoring, state estimation and control decisions. Traditional throughput and delay metrics are difficult to measure information freshness. To solve this problem, a new metric called Age of Information (AoI) was proposed and has received widespread attention.

With the continuous expansion of IoT deployment, the number of devices for status update services shows a massive growth trend. In such large-scale access scenarios, if the system continues to rely on centralized scheduling, it needs to maintain the states of a large number of nodes. This will incur extremely high control and feedback overhead. In contrast, random access avoids the need for global state information. It can support the dynamic access of large-scale devices with low complexity, making it more suitable for communication scenarios with massive devices. Therefore, how to balance access efficiency and information freshness under large-scale access conditions is an important MAC layer problem. Many scholars have conducted research on this problem, leading to numerous achievements in this field.

However, existing research does not fully consider the impact of continuous collisions in large-scale access scenarios. Moreover, most studies focus on the average AoI, paying less attention to tail metrics like the AoI violation probability. For time-sensitive services such as industrial control, considering only the average AoI is not enough to determine system reliability. To address these problems, in this thesis, we fully consider the negative impact of continuous collisions in large-scale access scenarios. First, we design a waiting-based random access backoff mechanism. Based on this, we build theoretical analysis models for two classic random access frameworks: slotted ALOHA and standard CSMA. Then, we derive closed-form expressions for the average AoI and AoI violation probability. The main contributions of this thesis are as follows:

\begin{enumerate} [wide]
\item \textbf{Designing a waiting mechanism for large-scale access scenarios.} To address the problem that existing research ignores continuous collisions in large-scale networks, we design a waiting-based random access backoff mechanism. This mechanism takes collision behaviors into account. By planning different backoff behaviors for nodes after successful updates and collisions, we effectively relieve excessive channel competition. This fills the gap of existing mechanisms in congestion handling.

\item \textbf{Introducing the proposed mechanism into the slotted ALOHA framework and analyzing its performance.} In large-scale networks, node states are coupled, making the system difficult to analyze. To solve this, we use a steady-state decoupling method to simplify the complex system into a single-node Markov chain model. Based on this, we establish and solve steady-state equations. Then, based on the obtained steady-state features, we derive closed-form expressions for the average AoI and AoI violation probability, and deeply analyze the impact of key system parameters on information timeliness.

\item \textbf{Introducing the proposed mechanism into the standard CSMA framework and analyzing its performance.} In large-scale networks, node states are coupled. Moreover, carrier sensing and backoff freezing mechanisms in the CSMA framework cause variable-length system slots. These two challenges make the system difficult to analyze. To solve this, we use a steady-state decoupling method to simplify the complex system into a single-node Markov chain model. Considering the variable-length slots, we establish and solve steady-state equations based on this model. Then, by combining the obtained steady-state features with the statistical characteristics of variable-length slots, we derive closed-form expressions for the average AoI and AoI violation probability, and deeply analyze the impact of key system parameters on information timeliness.

\item \textbf{Verifying the theoretical model and revealing the performance evolution rules of the waiting mechanism.} Through simulation experiments, we fully verify the accuracy of the derived closed-form expressions. Further comparative analysis shows that, under medium-to-high loads, especially in massive device access scenarios, moderate waiting does not harm information freshness. Instead, it reduces channel competition and lowers the collision probability, thereby improving both the average AoI and the AoI violation probability simultaneously.
\end{enumerate}

In summary, our research not only enriches the theoretical analysis of random access for Age of Information, but also provides valuable theoretical reference and engineering guidance for the protocol design and optimization of future large-scale IoT status update systems.

\end{abstractEN}

\begin{keywordEN}
Random access, Age of information, Performance analysis
\end{keywordEN}